---
title: "XAI3"
author: "G3"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The dataset is loaded and the necessary libraries for random forest modeling and PDP visualization are imported.
```{r, warning=FALSE}
library(randomForest)
library(pdp)
library(ggplot2)
library(readr)
library(dplyr)
library(patchwork)
data <- read_csv("model_data.csv", show_col_types = FALSE)
```

## 1. One dimensional Partial Dependence Plot. 
A random forest model is trained to predict bike rental counts using all available features.
```{r}
rf_model <- randomForest(cnt ~ ., data = data, ntree = 100, importance = TRUE)
```

The following row of partial dependence plots illustrates the individual effects of days since 2011, temperature, humidity, and wind speed on predicted bike rental counts.

```{r}
pdp_days <- partial(rf_model, pred.var = "days_since_2011", grid.resolution = 20)
plot_days <- autoplot(pdp_days) + ggtitle("Days since 2011")

pdp_temp <- partial(rf_model, pred.var = "temp_real", grid.resolution = 20)
plot_temp <- autoplot(pdp_temp) + ggtitle("Temperature")

pdp_humidity <- partial(rf_model, pred.var = "hum_real", grid.resolution = 20)
plot_humidity <- autoplot(pdp_humidity) + ggtitle("Humidity")

pdp_windspeed <- partial(rf_model, pred.var = "windspeed_real", grid.resolution = 20)
plot_windspeed <- autoplot(pdp_windspeed) + ggtitle("Wind Speed")

plot_days + plot_temp + plot_humidity + plot_windspeed
```

## 2. Bidimensional Partial Dependency Plot. 
A random sample of 50 rows is extracted from the dataset to generate a 2D Partial Dependence Plot showing how humidity and temperature affect bike rental predictions.
```{r}
set.seed(123)

sample_data <- data %>% sample_n(50)

rf_sample <- randomForest(cnt ~ ., data = sample_data, ntree = 100)

pdp_2d <- partial(rf_sample,
                  pred.var = c("hum_real", "temp_real"),
                  grid.resolution = 20,
                  progress = "none")

ggplot(pdp_2d, aes(x = temp_real, y = hum_real, fill = yhat)) +
  geom_tile(width = 1, height = 1) +
  scale_fill_viridis_c() +
  labs(title = "2D Partial Dependence Plot",
       x = "Temperature",
       y = "Humidity",
       fill = "Predicted Bike Rentals") +
  theme_minimal()
```

